---
title: "In Class Activity - Regression"
date: "16th Feb, 2022"
output: html_notebook
---

# Question 1. 

Data: ‘fev.csv’

```{r}
data <- read.csv("fev.csv")
str(data)
summary(data)
```


## 1.1. Fit a linear regression model with fev as response and the smoke indicator as predictor. Assess the assumptions of linearity, constant variance and normality using residual plots
```{r}
regrn <- lm(fev ~ factor(smoke), data = data)
plot(regrn)
```

## 1.2. Linearity?
Yes

## 1.3. Constant variance?
Yes

## 1.4. Normality?
Yes (somewhat not at tails, but large sample!)

## 1.5. What is your overall conclusion about validity of the results from the linear regression model?
Assumptions are largely met. 

## 1.6. Add age as a predictor variable to the model and create the residual plots for this model. How do they compare with the residual plots from the first model?
```{r}
plot(lm(fev ~ age + smoke, data=data))
```


# Question 2.

Data: ‘field_goals_sample.csv’

The data are from a random sample of 100 field goal attempts in National Football League games from 2008.

The variables to be used for this exercise are:

distance: distance from ball placement to goal posts (yards)
GOOD: indicator of whether the attempt was successful

```{r}
fg.data <- read.csv("field_goals_sample.csv")
str(fg.data)
summary(fg.data)
```


## 2.1. Compare the proportions of field goals that are successful for attempts at distance 40 yards or less versus more than 40 yards using the large-sample Z-test with significance level 0.05. Report the p-value from the test.
```{r}
library(BSDA)
library(dplyr)
le.40 <- fg.data %>% filter(distance <= 40)
gt.40 <- fg.data %>% filter(distance > 40)

BSDA::z.test(le.40$GOOD, gt.40$GOOD, sigma.x = sd(le.40$GOOD), sigma.y = sd(gt.40$GOOD))
```

## 2.2. Perform the test using the chi-square test with and without continuity correction.

## 2.3. Perform the test using Fisher’s exact test.

## 2.4. How do the results of the different tests compare?

## 2.5. Fit a logistic regression model to the data with outcome variable GOOD and predictor variable equal to an indicator that the distance is equal to 40 yards or less. 

## 2.6. Calculate the odds-ratio and state its interpretation.

## 2.7. Perform a test of the null hypothesis of that the coefficient of the indicator of distance being 40 or less is equal to 0.

## 2.8. How does this test result compare to the results of the previous tests?

## 2.9. Fit a logistic regression model to the data with predictor variable equal to the distance.

## 2.10. Calculate the odds-ratio for this model and state its interpretation.

## 2.11. Test the null hypothesis of no association between GOOD and distance.

## 2.12. How does this test result compare with that from the previous logistic regression model?
